Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-04-02 manuscript.qmd
refactor shared code & add to formalism

@@

However, we did not consider this straightforward approach, as it lacks a way to describe the correlation between ipsi- and contralateral involvement. This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.

```{python}
#| echo: false
#| label: tbl-contra
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.

import pandas as pd
import numpy as np
from scripts import paths, COL, CONTRA_LNLS

def right_align(column) -> str:
  """Right align some columns."""
  return ["text-align: right"] * len(column)

data = pd.read_csv(paths.filtered, header=[0,1,2])
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

@@

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized (state $\centernot{\text{e}}$) and every time-step there is a finite probability $\epsilon$ that the tumor grows over the symmetry plane of the patient (into state $\text{e}$).

@@

As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\pi}_\text{e} = \begin{pmatrix} 1 & 0 \end{pmatrix}$:

$$
\boldsymbol{\xi}^\text{e}[t] = \boldsymbol{\pi}_\text{e} \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t
$$

@@

@@

```{python}
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
#| echo: false

import yaml

import pandas as pd
from emcee.backends import HDFBackend
from lyscripts.utils import create_model

with open("models/midline/params.yaml", mode="r") as file:
  config = yaml.safe_load(file)

model = create_model(config)
backend = HDFBackend("models/midline/samples.hdf5", read_only=True)
chain = backend.get_chain(flat=True)

param_names = list(model.get_params().keys())
param_means = chain.mean(axis=0)
param_stds = chain.std(axis=0)
model.set_params(*param_means)
early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
late_midext_prob = model.get_distribution("late").pmf @ model.midext_evo()[:,1]

params_table = pd.DataFrame({
  "name": param_names,
  "mean": param_means,
  "std": param_stds,
})
(
  params_table.style
  .format("{:.2%}", subset=["mean", "std"])
  .apply(right_align, subset=["mean", "std"])
  .hide()
)
```

>>>>

```{python}
#| echo: false

from lymph import models
from lyscripts import utils

from scripts import shared, paths, COL, CONTRA_LNLS
```

@@

However, this approach lacks a way to describe the correlation between ipsi- and contralateral involvement. This is displayed in @tbl-contra and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal line or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III.

```{python}
#| echo: false
#| label: tbl-contra
#| tbl-cap: Contralateral involvement depending on whether the primary tumor extends over the mid-sagittal line, the T-category, and whether the ipsilateral LNL III was involved or healthy.
def right_align(column) -> str:
  """Right align some columns."""
  return ["text-align: right"] * len(column)

data = pd.read_csv(paths.data, header=[0,1,2])
data[COL.t_stage] = data[COL.t_stage].apply(lambda x: "early" if x <= 2 else "late")
early_midext_data = (data.loc[data[COL.t_stage] == "early", COL.midext] == True).mean()
late_midext_data = (data.loc[data[COL.t_stage] == "late", COL.midext] == True).mean()
grouped = data.groupby(by=[COL.midext, COL.t_stage, COL.ipsi_III])

@@

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized and every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.

@@

As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\epsilon}[0] = \begin{pmatrix} 1 & 0 \end{pmatrix}$:

$$
\boldsymbol{\epsilon}[t] = \boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t = \begin{pmatrix} (1 - \epsilon)^t & 1 - (1 - \epsilon)^t \end{pmatrix}
$$

We can again collect the entire evolution in a matrix $\boldsymbol{\Lambda}_\epsilon$ that stores the probabilies $P(\boldsymbol{\epsilon} \mid t)$ for each time-step $t$:

$$
\boldsymbol{\Lambda}_\epsilon = P(\boldsymbol{\epsilon} \mid t) = \begin{pmatrix}
\boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^0 \\
\boldsymbol{\epsilon}[1] \cdot \mathbf{A}_\epsilon^1 \\
\vdots \\
\boldsymbol{\epsilon}[t] \cdot \mathbf{A}_\epsilon^t \\
\end{pmatrix}
$$

This quantity has $t$ rows, one for each time-step, and two columns: one for the decreasing probability of a lateralized tumor and one for the increasing probability of a tumor that has crossed the mid-sagittal line.

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The matrix $\boldsymbol{\Lambda}_\epsilon$ shown as heatmap of probabilities.
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import make_axes_locatable

from lyscripts.plot.utils import COLORS

nrows, ncols = 1, 1

model = shared.get_model(which="midline", load_samples=True)

fig, ax = plt.subplots(tight_layout=True)

cmap = LinearSegmentedColormap.from_list("redusz", [COLORS["gray"], COLORS["red"]])
im = ax.imshow(model.midext_evo().T, cmap=cmap)

# create an Axes on the right side of ax. The width of cax will be 5%
# of ax and the padding between cax and ax will be fixed at 0.05 inch.
divider = make_axes_locatable(ax);
cax = divider.append_axes("right", size="5%", pad=0.05);
plt.colorbar(im, cax=cax, label="Prob. of midext");

ax.set_xticks(np.arange(model.max_time + 1));
ax.set_xlabel("time-step $t$");

ax.set_yticks([0., 1.], labels=["lat.", "ext."])
```

@@

## Prevalence Comparisons

Below we compare the predicted with the observed prevalence of some involvement patterns.

```{python}
#| echo: false
#| label: fig-prevalences
#| fig-cap: Prevalences of different scenarios.
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import draw, BetaPosterior, Histogram, COLOR_CYCLE

model = shared.get_model(which="midline")
scenarios_config = utils.load_yaml_params("scenarios.yaml")
scenarios = Scenario.list_from_params(scenarios_config)

fig, ax = plt.subplots()

content = []
color = next(COLOR_CYCLE)
content += [
  Histogram.from_hdf5(
    filename=paths.get_filename(model="midline", kind="prevalences"),
    dataname=scenarios[0].md5_hash("prevalences"),
    color=color,
  ),
  BetaPosterior.from_hdf5(
    filename=paths.get_filename(model="midline", kind="prevalences"),
    dataname=scenarios[0].md5_hash("prevalences"),
  ),
]
color = next(COLOR_CYCLE)
content += [
  Histogram.from_hdf5(
    filename=paths.get_filename(model="midline", kind="prevalences"),
    dataname=scenarios[1].md5_hash("prevalences"),
    color=color,
  ),
  BetaPosterior.from_hdf5(
    filename=paths.get_filename(model="midline", kind="prevalences"),
    dataname=scenarios[1].md5_hash("prevalences"),
  ),
]
draw(
  axes=ax,
  contents=content,
  percent_lims=(2., 2.),
)
```

@@

```{python}
#| echo: false
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
model = shared.get_model(which="midline", load_samples=True)
samples = shared.get_samples(which="midline")

names = list(model.get_params().keys())
means, stds = samples.mean(axis=0), samples.std(axis=0)

early_midext_prob = model.get_distribution("early").pmf @ model.midext_evo()[:,1]
late_midext_prob = model.get_distribution("late").pmf @ model.midext_evo()[:,1]

params_table = pd.DataFrame({"name": names, "mean": means, "std": stds})
(
  params_table.style
  .format("{:.2%}", subset=["mean", "std"])
  .apply(right_align, subset=["mean", "std"])
  .hide()
)
```
